---
title: "Day 1- Solutions"
format: html
---

# Getting Started

## Installing Packages
```{r}
#| label: install-packages-1
#| eval: false
install.packages("tidyverse")
install.packages("fivethirtyeight")
install.packages("janitor")
install.packages("labelled")
```

OR 

```{r}
#| label: install-packages-2
#| eval: false
install.packages(c("tidyverse","fivethirtyeight",
                   "janitor","labelled"))
```

## Load Packages 

```{r}
#| label: load-packages
library(tidyverse)
library(fivethirtyeight)
library(janitor)
library(labelled)
```


# Data Visualization

## Load Data
```{r}
#| label: load-data-noecho
#| echo: false
edibnb <- readRDS("../data/edibnb.RDS")
```

```{r}
#| label: load-edibnb-data
#| eval: false
edibnb <- readRDS("PATHNAME/edibnb.RDS")
```

## Exercises
1. How many observations (rows) does the dataset have?
```{r} 
#| label: dv-nobs

nrow(edibnb)
```

2. Run `View(edibnb)` in your console to view the data in the data viewer. What does each row in the dataset represent?


3. Create a barchart:
    a. Displaying the count of listings in each neighborhood

```{r}
#| label: dv-3a
    ggplot(edibnb, aes(x = neighborhood)) +
      geom_bar()
```

    b. Add different colors for the number of reviews
```{r}
#| label: dv-3b
    ggplot(edibnb, aes(x = neighborhood, fill = number_of_reviews)) +
      geom_bar()
```

4. Create a visualization that will help you compare the distribution of review scores (`review_scores_rating`) across neighborhoods.
```{r}
#| label: dv-4
ggplot(edibnb, aes(x = review_scores_rating, color = neighborhood)) +
  geom_boxplot()

ggplot(edibnb, aes(x = review_scores_rating, fill = neighborhood)) +
  geom_histogram() +
  facet_wrap(~neighborhood) +
  theme_minimal()

ggplot(edibnb, aes(x = review_scores_rating, fill = neighborhood)) +
  geom_density(alpha = .5) +
  facet_wrap(~neighborhood)
```


# Tidy Data

Take your plot code and change it so it uses a pipe (`%>%`) workflow instead of calling the data directly in `ggplot()`

```{r}
#| label: tidy-sol
edibnb %>% 
  ggplot(aes(x = review_scores_rating, color = neighborhood)) +
  geom_boxplot()

edibnb %>% 
  ggplot(aes(x = review_scores_rating, fill = neighborhood)) +
  geom_histogram() +
  facet_wrap(~neighborhood) +
  theme_minimal()

edibnb %>% 
  ggplot(aes(x = review_scores_rating, fill = neighborhood)) +
  geom_density(alpha = .5) +
  facet_wrap(~neighborhood)
```

# Single Dataset - select, arrange, filter, distinct, count

## Review the data

```{r}
#| label: single-data
college_recent_grads
glimpse(college_recent_grads)
```

## Exercises

1. Which major has the highest percentage of women? (Hint: use `arrange(desc())` to easily see the top items).  Output only relevant variables

```{r}
#| label: single-1
college_recent_grads %>%
  arrange(desc(sharewomen)) %>%
  select(major, major_category, sharewomen)
```


2. Get a list of STEM majors that have median earnings < $36,000 (Hint: use the `%in%` to see if the major_category is in `stem_categories`).  Output only relevant variables

```{r}
#| label: single-2
stem_categories <- c(
  "Biology & Life Science",
  "Computers & Mathematics",
  "Engineering",
  "Physical Sciences"
)

stem_categories

college_recent_grads %>%
  filter(major_category %in% stem_categories & median < 36000) %>%
  select(major, major_category, median)
```

3. Output a count of the number of majors in each major_category

```{r}
#| label: single-3
college_recent_grads %>%
  count(major_category)
```

