---
title: "Day 2-Solutions"
format: html
---

# Single Dataset - mutate, group_by, summarize

## Load Packages & Set STEM Array
```{r}
#| label: single-prep
library(tidyverse)
library(fivethirtyeight)

stem_categories <- c("Biology & Life Science",
                         "Computers & Mathematics",
                         "Engineering",
                         "Physical Sciences")

stem_categories
```

## Exercises

1. Create a new variable that indicates if the major has more women than men

```{r}
#| label: single-1
more_women_flag <- college_recent_grads %>%
  mutate(more_women = if_else(sharewomen < .5, 0, 1)) %>%
  select(major, major_category, sharewomen, more_women)

more_women_flag %>%
  count(more_women)

more_women_flag %>%
  count(major_category, more_women)
```

2. Create a new variable called STEM that is a 0/1 indicator if the major is in one of the `stem_categories`.

```{r}
#| label: single-2
college_recent_grads_stem <- college_recent_grads %>%
  mutate(STEM = if_else(major_category %in% stem_categories, 1, 0))

college_recent_grads_stem %>%
  count(STEM)
```

3. Which major category has the highest typical (can use `mean()`, `median()`, etc.) median income? (Hint: use `group_by()` and `summarize()`)

```{r}
#| label: single-3
college_recent_grads %>%
  group_by(major_category) %>%
  summarize(mean_income = mean(median)) %>%
  arrange(desc(mean_income))

college_recent_grads %>%
  group_by(major_category) %>%
  summarize(median_income = median(median)) %>%
  arrange(desc(median_income))
```

4. Create a scatterplot of median income vs. proportion of women in that major, colored by whether the major is in a STEM field or not.

```{r}
#| label: single-4
library(scales)
college_recent_grads %>%
  mutate(STEM = if_else(major_category %in% stem_categories, "Yes", "No")) %>%
  ggplot(aes(x = sharewomen, y = median, fill = STEM)) +
  geom_point(color = "black", pch = 21, size = 3) +
  theme_minimal() +
  scale_fill_manual(name = "Stem Major", values = c("#FFA219", "#FECCDD")) +
  scale_y_continuous(labels = label_number()) +
  labs(
    x = "Proportion of women in major",
    y = "Median income of major"
  )
```


# Joining Datasets

## Install & Load Packages
```{r}
#| label: join-install
#| eval: false
install.packages("nycflights13")
```

```{r}
#| label: join-pkgs
library(nycflights13)
flights
weather
airports
airlines
planes
```

## Exercises

1. Which airline had the most Airbus planes departing NYC in 2013?

Option 1: Merge then do `case_when()`

```{r}
#| label: join-1-opt1

merged_planes<-flights %>% 
  select(tailnum, carrier) %>% 
  left_join(airlines, join_by(carrier)) %>% 
  left_join(planes, join_by(tailnum))


merged_planes %>% 
  select(tailnum, carrier, name, manufacturer) %>% 
  mutate(Airbus = case_when(manufacturer == "AIRBUS INDUSTRIE" ~ 1,
                            manufacturer == "AIRBUS" ~ 1,
                            TRUE ~ 0)) %>% 
  count(name, Airbus) %>% 
  filter(Airbus == 1)
```

Option 2: Merge then do `if_else()`

```{r}
#| label: join-1-opt2
merged_planes<-flights %>% 
  select(tailnum, carrier) %>% 
  left_join(airlines, join_by(carrier)) %>% 
  left_join(planes, join_by(tailnum))

merged_planes %>% 
  select(tailnum, carrier, name, manufacturer) %>% 
  mutate(Airbus = if_else((manufacturer == "AIRBUS INDUSTRIE" | 
                            manufacturer == "AIRBUS"),
                          1,0)) %>% 
  count(name, Airbus) %>% 
  filter(Airbus == 1)
```

Option 3: Merge then do `filter()`

```{r}
#| label: join-1-opt3
merged_planes<-flights %>% 
  select(tailnum, carrier) %>% 
  left_join(airlines, join_by(carrier)) %>% 
  left_join(planes, join_by(tailnum))

merged_planes %>% 
  select(tailnum, carrier, name, manufacturer) %>% 
  filter((manufacturer == "AIRBUS INDUSTRIE" | 
            manufacturer == "AIRBUS")) %>% 
  count(name)
```


Option 4: Can merge/filter in different orders.  Here we get the count by `carrier` then merge on the airline name.
```{r}
#| label: join-1-opt4

airbus_counts<-planes %>% 
  left_join(flights, join_by(tailnum)) %>% 
  select(tailnum, manufacturer, carrier) %>% 
  filter((manufacturer == "AIRBUS INDUSTRIE" | 
            manufacturer == "AIRBUS")) %>% 
  count(carrier) %>% 
  left_join(airlines, join_by(carrier)) %>% 
  select(name, n) %>% 
  arrange(name)

airbus_counts
```

2. What airlines do not have any Airbus planes? (Hint: use what you created in exercise 1 and use `anti_join()`)

```{r}
#| label: join-2

airlines %>% 
  anti_join(airbus_counts, join_by(name))
```


3. What airline had the most delayed over 30 min (based on arrival time)?

Option 1: can calculate a "Long Delay Variable"
```{r}
#| label: join-3-option1
flights %>%
  mutate(long_arr_delay = case_when(
    arr_delay > 30 ~ "Long Delay",
    TRUE ~ "Small or No Delay"
  )) %>%
  count(carrier, long_arr_delay) %>%
  filter(long_arr_delay == "Long Delay") %>%
  left_join(airlines, by = "carrier") %>%
  select(carrier, name, n) %>%
  arrange(desc(n))
```

Option 2: Can filter to the flights with delays over 30 min
```{r}
#| label: join-3-option2
flights %>%
  filter(arr_delay > 30) %>%
  count(carrier) %>%
  left_join(airlines, by = "carrier") %>%
  select(carrier, name, n) %>%
  arrange(desc(n))
```

4. Create a plot to see if there is a relationship between flight duration (`air_time`) and number of seats on the plane. (Hint: Only keep cases that have data in both datasets)

```{r}
#| label: join-4

flights %>% 
  select(tailnum, air_time) %>% 
  inner_join(planes, join_by(tailnum)) %>% 
  ggplot(aes(x = seats, y = air_time)) + 
  geom_point()
``` 

# Pivoting Data

## Data
We are using the same flight information as the exercises in the Joining Datasets section:
```{r}
#| label: pivot-install
#| eval: false
install.packages("nycflights13")
```

```{r}
#| label: pivot-pkgs
#| eval: false
library(nycflights13)
flights
weather
airports
airlines
planes
```

## Exercises
1. Create a crosstab output of the airline by departure airport (`origin`)

```{r}
#| label: pivot-1
flights %>% 
  count(carrier, origin) %>% 
  pivot_wider(names_from = "origin", values_from = "n")
```

With full names merged on:

**Notes:** 

    -  if the variable is called something different across the two datasets you can use `x_varname == y_varname` in the `join_by()` statement.
    -  if two datasets have the same name, the default suffix will be `.x` and `.y`.  You can specify a suffix, using a `suffix = c("_xsuffix", "_ysuffix")` statment in the `*_join()` statement

```{r}
#| label: pivot-1-labels
flights %>% 
  count(carrier, origin) %>% 
  left_join(airlines, join_by(carrier)) %>% 
  left_join(airports, join_by(origin == faa), suffix = c("_airline", "_airport")) %>% 
  select(n:name_airport) %>% 
  pivot_wider(names_from = name_airport, values_from = n) %>% 
  arrange(name_airline)
```

2. Calculate the average temperature, wind speed, and percipitation by origin airport for the month of January.  Try to do this using `pivot_longer()` and one `summarize()` statement.

```{r}
#| label: pivot-2
weather %>% 
  filter(month == 1) %>% 
  select(origin, temp, wind_speed, precip) %>% 
  pivot_longer(cols=temp:precip, 
               names_to = "weather_type",
               values_to = "weather_data") %>% 
  group_by(origin, weather_type) %>% 
  summarize(avg_weather_data = mean(weather_data, na.rm=TRUE)) %>% 
  pivot_wider(names_from = weather_type, 
              values_from = avg_weather_data)
```

3. Try to recreate this plot that displays the average wind speed and wind gusts by month on the same plot. (Hint: pivot to get the values and use `geom_line()`)

```{r}
#| label: pivot-3
weather %>% 
  select(month, wind_speed, wind_gust) %>% 
  pivot_longer(cols = wind_speed:wind_gust, 
               names_to="wind_type", 
               values_to="mph") %>% 
  group_by(month,wind_type) %>% 
  summarize(mph = mean(mph, na.rm=TRUE)) %>% 
  ggplot(aes(x = month, y = mph, color = wind_type)) + 
  geom_line() + 
  theme_bw()
```

