---
title: "Day 3-Solutions"
format: html
---

# Import/Export

## Data

All data for these exercises are located on our share drive.

## Exercises

1.  Read in the `svy_stata.dta` file and review the data structure with `glimpse()`.
2.  Convert all labelled data to factors and save as a new dataset called `stata_fcts`.
3.  Using the original data you read in, create a new dataset called `stata_nolabels` that removes all value labels
4.  Use `count()` to see the output differences between `Q2` on the original input data, `stata_fcts`, and `stata_nolabels`

# Data Attributes - Factors

## Data

Use `gss_cat` data from the `forcats` package. Remember you can get information on the built in datasets using `?gss_cat`.

```{r}
#| label: fct-pkgs
#| warning: false
library(tidyverse)
glimpse(gss_cat)
```

## Exercises

1.  Create a new age variable called `age_cat` that categorizes `age` into a factor with the following categories (Hint: Be mindful of any missing data):

    -   \< 30
    -   30-39
    -   40-49
    -   50-59
    -   60-69
    -   70 +

```{r}
#| label: fct-1
gss_cat_newage<-gss_cat %>% 
  mutate(age_cat = case_when(age < 30 ~ "< 30",
                             between(age, 30, 39) ~ "30-39",
                             between(age, 40, 49) ~ "40-49",
                             between(age, 50, 59) ~ "50-59",
                             between(age, 60, 69) ~ "60-69",
                             age >= 70 ~ "70+",
                             TRUE ~ "-99"),
         age_cat = factor(age_cat))

levels(gss_cat_newage$age_cat)

gss_cat_newage %>% 
  ggplot(aes(x = age_cat, y = age)) +
  geom_point()
```


2.  In the `relig` variable there are missing data not showing as `NA`.  Recode these to our usual -99 and -98 values to these options.

```{r}
#| label: fct-2
gss_cat_newrelig<-gss_cat %>% 
  mutate(relig2 = fct_recode(relig, 
                             "-99" = "No answer", 
                             "-98" = "Don't know"))

gss_cat_newrelig %>% 
  count(relig, relig2)
```


3. Create a new religion variable called `relig_other` that combines answers with \<100 responses into the `Other` category.  (Hint: Be mindful our the missing data you just recoded)

```{r}
#| label: fct-3

gss_cat_religoth<-gss_cat_newrelig %>% 
  mutate(relig_other = fct_other(relig2, 
                                 drop = c("Native american",
                                          "Other eastern",
                                          "Hinduism",
                                          "Orthodox-christian")))

gss_cat_religoth %>% 
  count(relig_other, relig2)
```

# Data Attributes - Date/Times

## Data

Use the `birthdays.csv` data that is hosted on the share drive.  Read in the data using `read_csv()`. (Hint: Remember to change pathnames to have forward slashes "/")
```{r}
#| label: date-data
#| warning: false
#| eval: false
birthdays<-read_csv("PATHNAME/birthdays.csv")
```

```{r}
#| label: date-data-read
#| echo: false
#| warnings: false

set.seed(20250715)
birthdays<-tibble(uniqueid = c(10000001:10000100),
                  birth_year=as.integer(runif(100, min=1920, max=2024)),
                  birth_month=as.integer(runif(100, min=1, max=13))) %>% 
  group_by(row_number()) %>% 
  mutate(birth_day = case_when(birth_month %in% c(4,6,9,11)~as.integer(runif(1,min=1,max=31)),
                         birth_month==2 & leap_year(birth_year)~as.integer(runif(1,min=1,max=30)),
                         birth_month==2~as.integer(runif(1,min=1,max=29)),
                         TRUE~as.integer(runif(1,min=1,max=32))),
         sex_prep = runif(1,min=0,max=1),
         sex = case_when(sex_prep < .45~"Male",
                         sex_prep < .9~"Female"),
         enrollment_year = as.integer(runif(1, min=birth_year+15, max=2024)),
         enrollment_month = case_when(!is.na(enrollment_year)~as.integer(runif(1, min=1, max=13))),
         enrollment_day = case_when(enrollment_month %in% c(4,6,9,11)~as.integer(runif(1,min=1,max=31)),
                                    enrollment_month==2 & leap_year(enrollment_year)~as.integer(runif(1,min=1,max=30)),
                                    enrollment_month==2~as.integer(runif(1,min=1,max=29)),
                                    !is.na(enrollment_month)~as.integer(runif(1,min=1,max=32)))) %>% 
  ungroup() %>% 
  select(uniqueid:birth_month,birth_day,sex:enrollment_day)
```

```{r}
#| label: date-glimpse
glimpse(birthdays)
```


## Exercises
1. Create a new birthday variable that is a date type variable.
```{r}
#| label: date-1
library(glue)
birthdays_dt<-birthdays %>% 
  mutate(birthday = ymd(glue("{birth_year} {birth_month} {birth_day}")))
birthdays_dt
```

2. Create a new entrollment variable that is a date type variable.

```{r}
#| label: date-2
birthdays_dt<-birthdays_dt %>% 
  mutate(enrollment_date = ymd(glue("{enrollment_year} {enrollment_month} {enrollment_day}")))
birthdays_dt
```

3. What is the average age of this population as of today? (Hint: use `today()` and `time_length()`)

```{r}
#| label: date-4
birthdays_dt %>% 
  mutate(age = time_length(today()-birthday,unit = "years")) %>% 
  summarize(avg_age = mean(age, na.rm=TRUE))
```

4. What is the average age of enrollment by sex? (Hint: use `time_length()`)

```{r}
#| label: date-3
birthdays_dt %>% 
  mutate(age_enrollment = time_length(enrollment_date-birthday,unit = "years")) %>% 
  group_by(sex) %>% 
  summarize(avg_age_enrollment = mean(age_enrollment, na.rm=TRUE))
```

# Data Attributes - Strings

## Data

Use the `gss_cat` dataset from the `forcats` package

## Exercises
1. Create a new variable for party affiliation that creates 4 categories.  Use stringr functions to find these instead of manually typing all.

    - Independent
    - Democrat (include strong, not strong, and near)
    - Republican (include strong, not strong, and near)
    - Other (include strong, not strong, and near)

```{r}
#| label: string-1

gss_cat_party<-gss_cat %>% 
  mutate(party_affil = case_when(str_detect(tolower(partyid), "dem") ~ "Democrat",
                                 str_detect(tolower(partyid), "rep") ~ "Republican",
                                 TRUE ~ partyid)) 

gss_cat_party %>% 
  count(party_affil, partyid)

```

2. Create a new variable called `baptist` that indicates if a respondent is some type of baptist based on their response to `denom`.  (Hint: View the levels of `denom` to understand what might be considered baptist.)

```{r}
#| label: string-2

gss_cat_baptist<-gss_cat %>% 
  mutate(baptist = if_else(str_detect(tolower(denom), "bapt"), 1, 0)) 

gss_cat_baptist %>% 
  count(denom, baptist) %>%
  arrange(-baptist) %>% 
  print(n = 30)

```

3. The `rincome` variable has some levels using "to" for the wording (\$8000 to 9999) and some using "-" for the wording (\$10000 - 14999).  Create a new variable that standardizes this wording using the functions from stringr.

```{r}
#| label: string-3-opt1
gss_cat %>% 
  mutate(rincome_cat = str_replace(rincome," - ", " to ")) %>% 
  count(rincome, rincome_cat)
```

Option 2:
```{r}
#| label: string-3-opt2
gss_cat %>% 
  mutate(rincome_cat = str_replace(rincome," to ", " - ")) %>% 
  count(rincome, rincome_cat)
```
