---
title: "Day 4-Solutions"
format: html
---

# Data Attributes - Strings

## Data

Use the `gss_cat` dataset from the `forcats` package
```{r}
#| label: fct-pkgs
#| warning: false
library(tidyverse)
glimpse(gss_cat)
```

## Exercises
1. Create a new variable for party affiliation that creates 4 categories.  Use stringr functions to find these instead of manually typing all.

    - Independent
    - Democrat (include strong, not strong, and near)
    - Republican (include strong, not strong, and near)
    - Other (include strong, not strong, and near)

```{r}
#| label: string-1

gss_cat_party<-gss_cat %>% 
  mutate(party_affil = case_when(str_detect(tolower(partyid), "dem") ~ "Democrat",
                                 str_detect(tolower(partyid), "rep") ~ "Republican",
                                 TRUE ~ partyid)) 

gss_cat_party %>% 
  count(party_affil, partyid)

```

2. Create a new variable called `baptist` that indicates if a respondent is some type of baptist based on their response to `denom`.  (Hint: View the levels of `denom` to understand what might be considered baptist.)

```{r}
#| label: string-2

gss_cat_baptist<-gss_cat %>% 
  mutate(baptist = if_else(str_detect(tolower(denom), "bapt"), 1, 0)) 

gss_cat_baptist %>% 
  count(denom, baptist) %>%
  arrange(-baptist) %>% 
  print(n = 30)

```

3. The `rincome` variable has some levels using "to" for the wording (\$8000 to 9999) and some using "-" for the wording (\$10000 - 14999).  Create a new variable that standardizes this wording using the functions from stringr.

```{r}
#| label: string-3-opt1
gss_cat %>% 
  mutate(rincome_cat = str_replace(rincome," - ", " to ")) %>% 
  count(rincome, rincome_cat)
```

Option 2:
```{r}
#| label: string-3-opt2
gss_cat %>% 
  mutate(rincome_cat = str_replace(rincome," to ", " - ")) %>% 
  count(rincome, rincome_cat)
```

# Labels

Solutions are located on our sharedrive

