---
title: "Pivoting Data"
format: html
---

## Slides

[View slides in full screen](06_PivotingData.html)

```{=html}
<iframe class="slide-deck" src="06_PivotingData.html" height="450" width="800" style="border: 1px solid #F0F0F0;"></iframe>
```
## Exercises


## Data
We are using the same flight information as the exercises in the Joining Datasets section:
```{r}
#| label: pivot-install
#| eval: false
install.packages("nycflights13")
```

```{r}
#| label: pivot-pkgs
#| eval: false
library(nycflights13)
flights
weather
airports
airlines
planes
```

## Exercises
1. Create a crosstab output of the airline by depature airport (`origin`)
2. Calculate the average temperature, wind speed, and percipitation by origin airport for the month of January.  Try to do this using `pivot_longer()` and one `summarize()` statement.
3. Try to recreate this plot that displays the average wind speed and wind gusts by month on the same plot. (Hint: pivot to get the values Use `geom_line()`)

```{r}
#| label: pivot-3
#| echo: false
#| warning: false

library(tidyverse)
library(nycflights13)

weather %>% 
  select(month, wind_speed, wind_gust) %>% 
  pivot_longer(cols = wind_speed:wind_gust, 
               names_to="wind_type", 
               values_to="mph") %>% 
  group_by(month, wind_type) %>% 
  summarize(mph = mean(mph, na.rm=TRUE)) %>% 
  ggplot(aes(x = month, y = mph, color = wind_type)) + 
  geom_line() + 
  theme_bw()
```
  