---
title: "Working with a single dataset"
order: 4
title-slide-attributes:
    data-background-gradient: linear-gradient(to top right, {{< brand color background >}}, {{< brand color pastelpink >}}, {{< brand color fmorange >}}, {{< brand color tangerine >}})

format: 
  revealjs:
    margin: .15
    quarto-required: ">= 1.6.0"
    width: 1600
    height: 900
    slide-number: "c"
    css: ../styles.css
    code-copy: true
    echo: true
    code-overflow: scroll
---

```{r}
#| label: packages
#| echo: false
library(tidyverse)
library(emojifont)
options(width=69)
```


# Data: Penguins 

```{r}
#| label: penguin-prep
#| echo: false
penguins<-tibble(penguins)
penguins
```

# `select` and `arrange` {background-gradient="linear-gradient(to bottom right, #FFFFFF, #7CDCFF, #7CDCFF, #7CDCFF)"}

## `select` - keep variables

```{r}
#| label: select-keep
#| code-line-numbers: 2
penguins %>%
  select(species, island)
```

## `select` - exclude variables

```{r}
#| label: select-exclude
#| code-line-numbers: 2
penguins %>%
  select(-year)
```

## `select` - multiple variables

-   a range of variables

```{r}
#| label: select-range
#| code-line-numbers: 2
penguins %>%
  select(species:bill_dep)
```

## `select` - multiple variables

-   a range of variables
-   variables with certain characteristics

. . .

::: column
```{r}
#| label: select-starts
#| code-line-numbers: 2
penguins %>%
  select(starts_with("bill"))
```
:::

:::: column
::: fragment
```{r}
#| label: select-ends
#| code-line-numbers: 2
penguins %>%
  select(ends_with("len"))
```
:::
::::

## Select helpers

-   `starts_with()`: Starts with a prefix
-   `ends_with()`: Ends with a suffix
-   `contains()`: Contains a literal string
-   `num_range()`: Matches a numerical range like x01, x02, x03
-   `one_of()`: Matches variable names in a character vector
-   `everything()`: Matches all variables
-   `last_col()`: Select last variable, possibly with an offset
-   `matches()`: Matches a regular expression (a sequence of symbols/characters expressing a string/pattern to be searched for within text)

::: footnote
See help for any of these functions for more info, e.g. `?everything`.
:::

## `arrange` in ascending / descending order

```{r}
#| label: arrange-width
#| echo: false
options(width=32)
```

::: column
```{r}
#| label: arrange-ascending
#| code-line-numbers: 3
penguins %>%
  select(species,bill_dep) %>%
  arrange(bill_dep)
```
:::

::: column
```{r}
#| label: arrange-descending
#| code-line-numbers: 3
penguins %>%
  select(species,bill_dep) %>%
  arrange(desc(bill_dep))
```
:::

. . .

::: column
:::

::: column
```{r}
#| label: arrange-descending-minus
#| code-line-numbers: 3
penguins %>%
  select(species,bill_dep) %>%
  arrange(-bill_dep)
```
:::


# `filter` {background-gradient="linear-gradient(to bottom right, #FFFFFF, #FECCDD, #FECCDD, #FECCDD)"}

```{r}
#| label: filter-width
#| echo: false
options(width=69)
```

## `filter` to select a subset of rows

```{r}
#| labels: filter
#| code-line-numbers: 2
penguins %>%
  filter(species == "Adelie") 
```

## `filter` for many conditions at once

```{r}
#| labels: filter-multi
#| code-line-numbers: "3,4"
penguins %>%
  filter(
    species == "Adelie",
    bill_dep > 20
    ) %>%
  select(species:bill_dep)
```

## `filter` for more complex conditions

```{r}
#| labels: filter-complex
#| code-line-numbers: 4
penguins %>%
  filter(
    species == "Adelie",
    bill_dep > 20 | bill_dep < 17
    ) %>%
  select(species:bill_dep)
```

## Logical operators in R

<br>

| operator | definition |   | operator | definition |
|----------|------------------------|----|------------|------------------------|
| `<` | less than |  | `x & y` | `x` AND `y` |
| `<=` | less than or equal to |  |  `x | y` | `x` OR `y` |
| `>` | greater than |  | `is.na(x)` | test if `x` is `NA` |
| `>=` | greater than or equal to |  | `!is.na(x)` | test if `x` is not `NA` |
| `==` | exactly equal to |  | `x %in% y` | test if `x` is in `y` |
| `!=` | not equal to |  | `!(x %in% y)` | test if `x` is not in `y` |

# `distinct` and `count` {background-gradient="linear-gradient(to bottom right, #FFFFFF, #7CDCFF, #7CDCFF, #7CDCFF)"}


## `distinct` to filter for unique rows

... and `arrange` to order alphabetically

::: {.column}
```{r}
#| label: distinct-1var
#| code-line-numbers: 2
penguins %>%
  distinct(species) %>%
  arrange(species)
```
:::

:::: {.column}
::: fragment
```{r}
#| label: distinct-2var
#| code-line-numbers: 2
penguins %>%
  distinct(species, island) %>%
  arrange(species, island)
```
:::
::::

## `count` to create frequency tables

::: {.column}
Alphabetical ordering by default
<br>
```{r}
#| label: count-default
#| code-line-numbers: 2
penguins %>%
  count(species)
```
:::

:::: {.column}
::: fragment
Order by descending frequency
<br>
```{r}
#| label: count-descending-n
#| code-line-numbers: 2
penguins %>%
  count(species, sort = TRUE)

```
:::
::::

## `count` and `arrange`

::: {.column}
Arrange by ascending frequency order
```{r}
#| label: count-ascending-n
#| code-line-numbers: 3
penguins %>%
  count(species) %>%
  arrange(n)
```
:::

:::: {.column}
::: fragment
Arrange by descending frequency order
```{r}
#| label: count-descending-n2
#| code-line-numbers: 3
penguins %>%
  count(species) %>%
  arrange(desc(n))
```

*same output as `sort = TRUE`*
:::
::::

## `count` for multiple variables

```{r}
#| label: count-multi-1
#| code-line-numbers: 2
penguins %>%
  count(species,island) 
```

## order matters when you `count`

::: {.column}
`species` first
```{r}
#| label: count-multi-2
#| code-line-numbers: 2
penguins %>%
  count(species,island) 
```
:::

:::: {.column}
::: fragment
`island` first
```{r}
#| label: count-multi-3
#| code-line-numbers: 2
penguins %>%
  count(island,species) 
```
:::
::::

# Your turn: <br>`select`, `arrange`, `filter`, `distinct`, `count` {background-gradient="linear-gradient(to bottom, #FFFFFF, #FFFFFF, #6BE1CA)"}

`r emojifont::emoji("raised_hand")` if you have questions or need help

`r fontawesome::fa("comment")` when you are finished

```{r}
#| echo: false
countdown::countdown(minutes = 15, seconds = 0, play_sound = TRUE)
```


